ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs

### 2023-08-22 coin-or/main.tex
references added

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  % \includegraphics[width=.48\textwidth]{speedup-sol.pdf} 
  \label{fig:simd}
  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}\label{fig:memory}
\end{figure}

@@

Recently, we have discovered that the condensed-space interior point
method (IPM) with an inequality relaxation strategy is particularly
effective for solving large-scale NLPs, like AC optimal power flow, up
to moderate precisions. This strategy efficiently addresses the
challenges posed by sparse matrix factorization on GPUs, as the
condensed KKT systems can be factorized without numerical pivoting,
which has previously hindered GPU utilization in the large-scale
optimization regime.  Furthermore, MadNLP offers diverse approaches
for handling KKT systems with special structures, via reduction
strategies (citation needed) and Schur complement decomposition
strategies (citation needed).

@@

Our numerical results, summarized in Table \ref{tab:num} (for detailed
results, please refer to \cite{shin2023accelerating}), clearly
demonstrate the significant potential of using GPUs. Specifically,
when applied to AC OPF problems, which are one of the most important
applications of NLP, we have achieved significant performance
enhancements. The GPU-based solutions achieve speedups of over 10
times compared to CPU-based state-of-the-art tools. Notably, ExaModels
enables derivative evaluations that are up to two orders of magnitude
faster, and the use of the condensed-space IPM strategy in MadNLP
greatly improves linear algebra computation speed. Furthermore, we
have also explored the utilization of these methods in reduced-space
OPF \cite{} and model predictive control (MPC) \cite{}, where we have
achieved similar degrees of speedup.

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
  \end{center}
  \label{tab:num}
\end{table}

@@

\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
        \hline
        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
        \hline
        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the NLP frameworks.}
  \label{tbl:portabiltiy}
\end{table}

@@

\begin{itemize}
\item \textit{Energy Infrastructures}: Enhancing the scalability of
optimization methods is crucial for improving the resilience of energy
system operations. We have demonstrated the ability of MadNLP with
multiple GPUs to solve large-scale security-constrained AC OPF
problems effectively. This showcases the potential of ExaModels and
MadNLP to enhance the scalability of NLP by utilizing GPU
capabilities.

\item \textit{Model Predictive Control}: MPC has been widely used for
optimization-based control in various settings, such as chemical
processes, energy systems, and robotics. However, the effectiveness of
MPC is often limited by the efficiency of optimization solvers,
especially when short sampling times are required. Our results have
shown that MadNLP running on GPUs can solve such problems with high
efficiency \cite{}. Moving forward, we plan to implement the extension
packages for control problems so that control community can take
advantage of accelerated NLP frameworks.

\item \textit{Machine Learning Surrogate Models}: An exciting future
direction is the integration of neural network surrogate models into
constrained optimization algorithms. GPU acceleration plays a crucial
role in these applications, considering the complexity of real-world
AI models. MadNLP offers the necessary capabilities through the native
handling of device arrays as well as dense KKT systems with various
quasi-Newton strategies. We intend to explore these problems further
and provide extensions for surrogate modeling with GPU acceleration.

@@

\bibliography{main}
\end{document}

>>>>

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}
  \label{fig:simd}
\end{figure}

@@

Recently, we have discovered that the condensed-space interior point
method (IPM) with an inequality relaxation strategy is particularly
effective for solving large-scale NLPs, like AC optimal power flow, up
to moderate precisions. This strategy efficiently addresses the
challenges posed by sparse matrix factorization on GPUs, as the
condensed KKT systems can be factorized without numerical pivoting,
which has previously hindered GPU utilization in the large-scale
optimization regime.  Furthermore, MadNLP offers diverse approaches
for handling KKT systems with special structures, via reduction
strategies \cite{pacaud2023accelerating} and Schur complement decomposition
strategies \cite{pacaud2023parallel}.

@@

Our numerical results, summarized in Table \ref{tab:num} (for detailed
results, please refer to \cite{shin2023accelerating}), clearly
demonstrate the significant potential of using GPUs. Specifically,
when applied to AC OPF problems, which are one of the most important
applications of NLP, we have achieved significant performance
enhancements. The GPU-based solutions achieve speedups of over 10
times compared to CPU-based state-of-the-art tools. Notably, ExaModels
enables derivative evaluations that are up to two orders of magnitude
faster, and the use of the condensed-space IPM strategy in MadNLP
greatly improves linear algebra computation speed. Furthermore, we
have also explored the utilization of these methods in reduced-space
OPF \cite{pacaud2023accelerating} and model predictive control (MPC)
\cite{cole2023exploiting}, where we have achieved similar degrees of
speedup.

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  
  \end{center}
  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
  \label{tab:num}
\end{table}

@@

\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
        \hline
        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
        \hline
        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the NLP frameworks.}
  \label{tbl:portability}
\end{table}

@@

\begin{itemize}
\item \textit{Energy Infrastructures}: Enhancing the scalability of
optimization methods is crucial for improving the resilience of energy
system operations. We have demonstrated the ability of MadNLP with
multiple GPUs to solve large-scale security-constrained AC OPF
problems effectively \cite{pacaud2023parallel}. This showcases the
potential of ExaModels and MadNLP to enhance the scalability of NLP by
utilizing GPU capabilities.

\item \textit{Model Predictive Control}: MPC has been widely used for
optimization-based control in various settings, such as chemical
processes, energy systems, and robotics. However, the effectiveness of
MPC is often limited by the efficiency of optimization solvers,
especially when short sampling times are required. Our results have
shown that MadNLP running on GPUs can solve such problems with high
efficiency \cite{cole2023exploiting}. Moving forward, we plan to
implement the extension packages for control problems so that control
community can take advantage of accelerated NLP frameworks.

\item \textit{Machine Learning Surrogate Models}: An exciting future
direction is the integration of neural network surrogate models into
constrained optimization algorithms (a proof-of-concept presented in
\cite{shin2023constrained}). GPU acceleration plays a crucial role in
these applications, considering the complexity of real-world AI
models. MadNLP offers the necessary capabilities through the native
handling of device arrays as well as dense KKT systems with various
quasi-Newton strategies. We intend to explore these problems further
and provide extensions for surrogate modeling with GPU acceleration.

@@

\bibliographystyle{plain} 
\bibliography{main}
\end{document}
